\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{array}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{tikz}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[hidelinks]{hyperref}

\title{Lecture Notes on the IEEE 9-Bus AC Power-Flow Problem}
\author{Generated from the loadflow-lite MATPOWER fixtures}
\date{\today}

\newcommand{\pu}{\mathrm{p.u.}}
\newcommand{\MW}{\mathrm{MW}}
\newcommand{\MVAr}{\mathrm{MVAr}}

\begin{document}

\maketitle

\begin{abstract}
The IEEE 9-bus system is a compact benchmark for the steady-state AC
power-flow problem. It is small enough to inspect by hand, but it still
contains the main modeling features that make power flow nontrivial:
multiple generator buses, voltage magnitude control, nonlinear active-
and reactive-power balance equations, branch losses, and reactive-power
exchange with transmission lines. These notes formulate the problem,
describe the specific 9-bus network used in this repository, discuss
the solved operating point recorded in the checked-in MATPOWER-derived
fixtures, and offer an economic interpretation for readers whose primary
background is markets rather than power engineering.
\end{abstract}

\section{Context and Relevance}

Power-flow analysis is the basic steady-state calculation behind many
power-system engineering tasks. Given a network model and specified
generation, load, and voltage-control data, the AC power-flow problem
computes the complex voltage at each bus and the real and reactive power
flows on each branch. The result is not a dynamic trajectory; it is a
snapshot of an operating point under balanced, sinusoidal, three-phase
conditions.

This calculation matters because many operational and planning decisions
depend on quantities that are not specified directly in the input data:
voltage magnitudes at load buses, bus voltage angles, slack-bus real-power
output, generator reactive-power output, branch flows, and system losses.
A candidate dispatch may satisfy a real-power energy balance in aggregate
while still producing unacceptable voltage profiles, excessive line flows,
or reactive-power requirements. AC power flow is therefore the entry point
for contingency analysis, voltage-security assessment, optimal power flow,
state estimation validation, and many simulation studies.

It is also the physical feasibility model behind many nodal and
security-constrained wholesale market calculations. Locational marginal
prices (LMPs) are the dual variables on nodal active-power balance
constraints in optimal power flow. Multipliers on binding branch-flow
constraints produce congestion components of LMPs and settlement
congestion rents, but the rents are not themselves dual variables. The
slack-bus convention fixes the angle reference. Many production-grade
market engines use a linearized \emph{DC} power-flow approximation of
the equations developed here, while AC power flow is the more complete
network model those approximations are checked against. Understanding
the AC equations on a small, transparent case is therefore useful
background for anyone working with electricity-market models.

The case used here is the WSCC 3-machine 9-bus system, often called the
Anderson 9-bus after its appearance in Anderson and Fouad's
\emph{Power System Control and Stability}. It is useful pedagogically
because it is small but not degenerate. It has three generator buses
(one slack, two PV), six PQ buses, and nine branches; impedances are
reported on a 100 MVA system base. The topology is a meshed transmission
network rather than a radial feeder, and the solution exhibits both
real-power losses and significant reactive-power effects from line
charging.

\section{A Glossary for Readers from Economics}\label{sec:glossary}

This note assumes familiarity with linear algebra and Newton's method but
not with power engineering. The vocabulary is the main barrier. The
following one-line glosses cover the terms that appear repeatedly. An
economic analogue is given where one is genuinely close.

\begin{description}
  \item[Bus.] A node in the network --- physically, a substation
  busbar. Read as ``node'' throughout.
  \item[Branch.] An edge between two buses. May be a transmission line or
  a transformer. Read as ``arc'' or ``link.''
  \item[Per-unit (p.u.).] A non-dimensionalization. Every quantity is
  divided by a reference base value, so a voltage of $1.00~\pu$ means
  ``the nominal design voltage at this bus,'' and a power of $1.00~\pu$
  means $100~\MW$ on the system used here (100~MVA base). Treat it
  exactly like reporting macro variables as fractions of GDP --- the
  base is sometimes implicit but always there.
  \item[Real power (MW).] The energy actually delivered per unit time;
  what consumers pay for and what generators are dispatched on.
  \item[Reactive power (MVAr).] A network-support quantity needed to
  sustain AC voltages in the presence of inductive and capacitive
  elements. It does not deliver useful work in the same sense as real
  power, but transmission cannot operate without it. Market treatment
  varies by jurisdiction: reactive support may be covered by
  interconnection requirements, cost-based compensation, tariff
  obligations, or ancillary-service procurement rather than a nodal price
  analogous to LMP.
  \item[Slack bus.] The bus where voltage magnitude and angle are fixed
  by convention and the generator output is whatever the rest of the
  problem implies. It is analogous to a numéraire only in the narrow
  sense that one reference is fixed: here one angle is fixed at zero, and
  all other angles are then \emph{relative} to it.
  \item[PV bus.] A generator bus with specified real-power output and
  voltage magnitude (the ``P'' and the ``V''). Reactive output and angle
  are solved. In this fixture, reactive limits are not enforced, so the
  solved $Q_G$ is whatever the unconstrained power-flow equations require
  to hold the voltage setpoint.
  \item[PQ bus.] A bus with specified real and reactive demand (the ``P''
  and the ``Q''). Both voltage magnitude and angle are solved. Read as
  ``a passive consumption node.''
  \item[Voltage angle ($\theta_i$).] A scalar at each node. Differences in
  $\theta$ across a branch drive real-power flow on that branch. The
  closest economic analogy is to a relative-gradient variable: angle
  differences, like price differences, help determine the direction and
  quantity of exchange, but they are physical state variables rather than
  market prices.
  \item[Series impedance.] The branch parameters $r$ and $x$. Resistance
  $r$ produces real-power losses (the ``transaction cost''
  interpretation); reactance $x$ governs how strongly an angle
  difference pushes real power across the branch.
  \item[Shunt charging ($b$).] The total branch charging susceptance in
  the MATPOWER branch model. The model places $jb/2$ at each end of the
  branch; for positive capacitive charging, this contributes reactive
  injection that depends on the local voltage magnitude.
  \item[Tap ratio, phase shift.] Transformer settings. Both equal $1$
  and $0$ respectively in this fixture, so they can be ignored here.
  \item[Generator step-up transformer (GSU).] The transformer that
  connects a generator to the high-voltage transmission grid. The IEEE
  9-bus has three of them (branches 1--4, 3--6, 8--2).
\end{description}

A useful one-liner, with the analogy kept deliberately loose:
\emph{angle differences drive real-power exchange, real power is the
commodity, reactive power is voltage support, losses are the transaction
cost, and the slack bus supplies the reference angle.}

\section{Mathematical Formulation}

\subsection{Network Model}

Let the system have $n$ buses and complex bus-voltage phasors
\[
  V_i = |V_i| e^{j\theta_i}, \qquad i = 1,\ldots,n.
\]
The nodal admittance matrix is
\[
  Y_{\mathrm{bus}} = G + jB,
\]
where $G_{ik}$ and $B_{ik}$ are the conductance and susceptance entries
of the bus admittance matrix. The series-admittance part is
Laplacian-like, with weights set by branch impedances; line charging,
shunts, transformer taps, and phase shifts add the details that make the
full $Y_{\mathrm{bus}}$ more than a plain graph Laplacian. At each bus,
the net complex power injection is
\[
  S_i = P_i + jQ_i = V_i I_i^*,
\]
where $P_i$ is real power and $Q_i$ is reactive power. Using
$I = Y_{\mathrm{bus}}V$, the standard polar-form AC power-flow equations
are
\begin{align}
  P_i &=
  |V_i| \sum_{k=1}^n |V_k|
  \left[
    G_{ik}\cos(\theta_i-\theta_k)
    + B_{ik}\sin(\theta_i-\theta_k)
  \right], \\
  Q_i &=
  |V_i| \sum_{k=1}^n |V_k|
  \left[
    G_{ik}\sin(\theta_i-\theta_k)
    - B_{ik}\cos(\theta_i-\theta_k)
  \right].
\end{align}

The specified net injections are formed from positive generation and
positive load:
\[
  P_i^{\mathrm{spec}} = P_{G,i} - P_{D,i}, \qquad
  Q_i^{\mathrm{spec}} = Q_{G,i} - Q_{D,i}.
\]
All real and reactive powers in these notes are reported on the 100 MVA
base used by the fixture files unless explicitly shown in MW or MVAr.

\subsection{Bus Types and Unknowns}

The power-flow equations are solved with different specifications at
different bus types:
\begin{itemize}
  \item \textbf{Slack (reference) bus}: specifies $|V|$ and $\theta$. Its
  real and reactive generation are outputs of the solution. The slack
  bus serves two structural roles. First, it provides the angle
  reference $\theta_1 = 0$, without which the system is invariant under a
  global phase shift and the equations are underdetermined --- the
  reference-angle role. Second, its real-power output absorbs the
  active-power residual caused by losses once the non-slack real-power
  injections are fixed. Reactive power is handled differently: the slack
  bus and all PV buses have solved reactive outputs, so reactive balance
  is not assigned to a single bus.
  \item \textbf{PV bus}: specifies real-power generation $P_G$ and voltage
  magnitude $|V|$. The voltage angle $\theta$ and the reactive generation
  $Q_G$ are both solved. It is therefore wrong to read the case-side
  $Q_G$ at a PV bus as a setpoint; in the fixtures it is just an initial
  value that the solver overwrites.
  \item \textbf{PQ bus}: specifies real and reactive load. Both $|V|$ and
  $\theta$ are solved.
\end{itemize}

The resulting nonlinear system is square: there are $n - 1$ active-power
equations (one per non-slack bus) and $n - n_{\mathrm{PV}} - 1$
reactive-power equations (one per PQ bus). It is solved by Newton's
method on the mismatch residual; the standard derivation, including the
four Jacobian sub-blocks, is sketched in Appendix~\ref{app:newton}.

\section{The Specific IEEE 9-Bus Network}

The fixture used here is \texttt{data/case9.json}, with solved values from
\texttt{data/case9\_solution.json}. The case has three generator buses:
bus 1 is the slack bus, while buses 2 and 3 are PV buses. Loads are
located at buses 5, 7, and 9. The total load is
\[
  P_D = 315.00~\MW, \qquad Q_D = 115.00~\MVAr.
\]
The case-side $P_G$ at the slack bus and $Q_G$ at all generator buses
are MATPOWER initial values rather than scheduled targets, so it is the
solved generator outputs that should be compared against the load. Only
$P_G$ at the two PV buses (163.00 and 85.00~MW) is a true setpoint, and
the solver honors both exactly.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  busnode/.style={circle, draw, minimum size=7mm, inner sep=0pt, font=\small, fill=white},
  slackbus/.style={busnode, double, double distance=1pt},
  pvbus/.style={busnode, thick},
  loadbus/.style={busnode},
  branch/.style={thick},
  xfmr/.style={thick, dashed},
  scale=1.05
]

\node[slackbus] (b1) at (5.0, 8.8) {1};
\node[pvbus]    (b2) at (1.2, 2.2) {2};
\node[pvbus]    (b3) at (8.8, 2.2) {3};
\node[busnode]  (b4) at (5.0, 7.0) {4};
\node[loadbus]  (b5) at (2.6, 5.7) {5};
\node[busnode]  (b6) at (2.6, 3.6) {6};
\node[loadbus]  (b7) at (5.0, 2.4) {7};
\node[busnode]  (b8) at (7.4, 3.6) {8};
\node[loadbus]  (b9) at (7.4, 5.7) {9};

\draw[xfmr]   (b1) -- (b4);
\draw[branch] (b4) -- (b5);
\draw[branch] (b5) -- (b6);
\draw[xfmr]   (b3) -- (b6);
\draw[branch] (b6) -- (b7);
\draw[branch] (b7) -- (b8);
\draw[xfmr]   (b8) -- (b2);
\draw[branch] (b8) -- (b9);
\draw[branch] (b9) -- (b4);

\node[font=\scriptsize, anchor=west] at (5.2, 8.8) {slack};
\node[font=\scriptsize, anchor=east] at (1.0, 2.2) {PV};
\node[font=\scriptsize, anchor=west] at (9.0, 2.2) {PV};
\node[font=\scriptsize, anchor=east] at (2.4, 5.7) {load};
\node[font=\scriptsize, anchor=north] at (5.0, 2.1) {load};
\node[font=\scriptsize, anchor=west] at (7.6, 5.7) {load};

\end{tikzpicture}
\caption{Topology of the IEEE 9-bus (WSCC) system. The double-circle
node is the slack bus; thick-circle nodes are PV generator buses;
``load'' annotations mark the three PQ buses with non-zero demand. Dashed
lines (1--4, 3--6, 8--2) are the generator step-up transformer branches;
solid lines are transmission branches. The interior loop
$4{-}5{-}6{-}7{-}8{-}9{-}4$ is the meshed sub-network through which
generator power redistributes.}
\end{figure}

\begin{table}[htbp]
\centering
\caption{Bus data and solved voltages and generation for the IEEE 9-bus
case.}
\small
\setlength{\tabcolsep}{3pt}
\begin{tabular}{@{}rcrrrrrrrr@{}}
\toprule
Bus & Type & $P_D$ & $Q_D$ & $P_G^{\mathrm{case}}$ & $Q_G^{\mathrm{case}}$ &
$|V|$ & $\theta$ & $P_G^{\mathrm{sol}}$ & $Q_G^{\mathrm{sol}}$ \\
 & & MW & MVAr & MW & MVAr & p.u. & deg & MW & MVAr \\
\midrule
1 & slack & 0.00 & 0.00 & 72.30 & 27.03 & 1.0400 & 0.0000 & 71.64 & 27.05 \\
2 & PV & 0.00 & 0.00 & 163.00 & 6.54 & 1.0250 & 9.2800 & 163.00 & 6.65 \\
3 & PV & 0.00 & 0.00 & 85.00 & -10.95 & 1.0250 & 4.6648 & 85.00 & -10.86 \\
4 & PQ & 0.00 & 0.00 & 0.00 & 0.00 & 1.0258 & -2.2168 & 0.00 & 0.00 \\
5 & PQ & 90.00 & 30.00 & 0.00 & 0.00 & 1.0127 & -3.6874 & 0.00 & 0.00 \\
6 & PQ & 0.00 & 0.00 & 0.00 & 0.00 & 1.0324 & 1.9667 & 0.00 & 0.00 \\
7 & PQ & 100.00 & 35.00 & 0.00 & 0.00 & 1.0159 & 0.7275 & 0.00 & 0.00 \\
8 & PQ & 0.00 & 0.00 & 0.00 & 0.00 & 1.0258 & 3.7197 & 0.00 & 0.00 \\
9 & PQ & 125.00 & 50.00 & 0.00 & 0.00 & 0.9956 & -3.9888 & 0.00 & 0.00 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[htbp]
\centering
\caption{Branch parameters for the 9-bus case. All series and charging
parameters are in p.u. on the system base; \texttt{tap\_ratio} = 1.0 and
\texttt{phase\_shift} = 0 for every branch in this fixture, so those
columns are omitted. Branches with $r=0$ and $b=0$ (1--4, 3--6, 8--2) are
the generator step-up transformers in the original WSCC model; the
fixture stores them as plain reactances because per-unit base voltages
are not tracked here.}
\begin{tabular}{@{}rrrrrr@{}}
\toprule
Branch & From & To & $r$ & $x$ & $b$ \\
\midrule
1 & 1 & 4 & 0.00000 & 0.05760 & 0.00000 \\
2 & 4 & 5 & 0.01700 & 0.09200 & 0.15800 \\
3 & 5 & 6 & 0.03900 & 0.17000 & 0.35800 \\
4 & 3 & 6 & 0.00000 & 0.05860 & 0.00000 \\
5 & 6 & 7 & 0.01190 & 0.10080 & 0.20900 \\
6 & 7 & 8 & 0.00850 & 0.07200 & 0.14900 \\
7 & 8 & 2 & 0.00000 & 0.06250 & 0.00000 \\
8 & 8 & 9 & 0.03200 & 0.16100 & 0.30600 \\
9 & 9 & 4 & 0.01000 & 0.08500 & 0.17600 \\
\bottomrule
\end{tabular}
\end{table}

Topologically, the meshed sub-network is the loop
\[
  4 \rightarrow 5 \rightarrow 6 \rightarrow 7 \rightarrow 8
  \rightarrow 9 \rightarrow 4.
\]
Each generator feeds into this loop through a single radial branch:
gen 1 via branch 1--4, gen 2 via branch 8--2, and gen 3 via branch 3--6.
On the radial hops there is no choice of path, but once power enters the
loop it divides according to the network impedances and the bus
voltage-angle profile. This is the structural reason the angle profile
matters: it is the state that determines how much real power takes each
direction around the loop.

\section{Discussion of the Solved Operating Point}

\subsection{Active-power balance and losses}

The solved generator outputs are
\[
\begin{aligned}
  P_G^{\mathrm{sol}} &= 319.641~\MW, \\
  Q_G^{\mathrm{sol}} &= 22.840~\MVAr.
\end{aligned}
\]
Since the total real load is $315.000~\MW$, the real-power loss is
\[
  P_{\mathrm{loss}} = 319.641 - 315.000 = 4.641~\MW,
\]
which equals the sum of $P_f + P_t$ over all branches as it must. The
slack bus supplies $71.641~\MW$; the PV buses hold their scheduled
real-power outputs of $163.000~\MW$ at bus 2 and $85.000~\MW$ at bus 3.

\subsection{Voltage profile}

The voltage magnitudes remain close to nominal. The highest solved
voltage is the slack-bus voltage, $1.0400~\pu$. The lowest voltage is at
bus 9, $0.9956~\pu$, which is also the largest load bus
($125~\MW$ and $50~\MVAr$). The drop is driven primarily by the local
reactive demand: increasing reactive load $Q_D$ tends to lower voltage
at heavily loaded buses, while increasing net reactive injection tends
to raise it. Bus 9 has both the largest $Q_D$ and two relatively
impedance-heavy branches connecting it to the rest of the network.

A subtler observation is that bus 6 sits at $1.0324~\pu$, which is
\emph{above} the $1.0250~\pu$ PV setpoint used at generator buses 2 and
3. Bus 6 is directly adjacent to generator bus 3, and it also sits in a
meshed region with non-negligible line charging and nearby voltage
control. The solved branch-end quantities show mixed local reactive
exchange at bus 6, so the safest interpretation is not that one line
``causes'' the higher voltage by itself, but that the network solution
allows an interior PQ-bus voltage to rise above a generator voltage
setpoint. This is why the PV-setpoint values cannot be read as upper
bounds on the system voltage profile.

\subsection{Angles and active-power flow}

The voltage angles span from $-3.9888^\circ$ at bus 9 to $9.2800^\circ$
at bus 2. In an AC transmission network, active-power transfer is
strongly tied to angle differences across each branch through the
near-lossless approximation
\[
  P_{ij} \;\approx\; \frac{|V_i|\,|V_j|}{x_{ij}}\,\sin(\theta_i - \theta_j).
\]
Branch 8--2 illustrates this clearly. The fixture stores the orientation
as 8 to 2, with $P_f = -163.000~\MW$ and $P_t = 163.000~\MW$, so the
real-power flow is from bus 2 toward bus 8. Plugging in
$|V_2| = 1.0250$, $|V_8| = 1.0258$, $x_{8\text{-}2} = 0.0625$, and
$\theta_2 - \theta_8 = 9.2800^\circ - 3.7197^\circ = 5.5603^\circ$:
\[
  P_{28} \;\approx\; \frac{1.0250 \times 1.0258}{0.0625}\,
  \sin(5.5603^\circ) \;=\; 1.6303~\pu \;=\; 163.03~\MW,
\]
which matches the solved $163.00~\MW$ to three figures. The lossless
approximation is essentially exact on this branch because $r = 0$.
Reading the table this way --- always check angle differences when
interpreting branch flows --- is the most useful intuition the
9-bus case offers.

\subsection{Reactive-power balance}

\begin{longtable}{@{}rrrrrrrr@{}}
\caption{Solved branch powers. The ``from'' and ``to'' quantities follow
the fixture branch orientation. Positive endpoint power is injected from
that bus terminal into the branch; negative endpoint power is delivered
from the branch into that bus terminal.}\label{tab:branch-flows}\\
\toprule
From & To & $P_f$ & $Q_f$ & $P_t$ & $Q_t$ & $P_{\ell}$ & $Q_{\ell}$ \\
 & & MW & MVAr & MW & MVAr & MW & MVAr \\
\midrule
\endfirsthead
\toprule
From & To & $P_f$ & $Q_f$ & $P_t$ & $Q_t$ & $P_{\ell}$ & $Q_{\ell}$ \\
 & & MW & MVAr & MW & MVAr & MW & MVAr \\
\midrule
\endhead
1 & 4 & 71.641 & 27.046 & -71.641 & -23.923 & 0.000 & 3.123 \\
4 & 5 & 30.704 & 1.030 & -30.537 & -16.543 & 0.166 & -15.513 \\
5 & 6 & -59.463 & -13.457 & 60.817 & -18.075 & 1.354 & -31.531 \\
3 & 6 & 85.000 & -10.860 & -85.000 & 14.955 & 0.000 & 4.096 \\
6 & 7 & 24.183 & 3.120 & -24.095 & -24.296 & 0.088 & -21.176 \\
7 & 8 & -75.905 & -10.704 & 76.380 & -0.797 & 0.475 & -11.502 \\
8 & 2 & -163.000 & 9.178 & 163.000 & 6.654 & 0.000 & 15.832 \\
8 & 9 & 86.620 & -8.381 & -84.320 & -11.313 & 2.300 & -19.694 \\
9 & 4 & -40.680 & -38.687 & 40.937 & 22.893 & 0.258 & -15.794 \\
\bottomrule
\end{longtable}

The sign discussion generalizes: on branch $7$--$8$, the negative $P_f$
indicates that real power enters the branch at bus 8 and leaves at bus
7, even though the fixture orientation runs the other way.

The reactive-power column tells a different story. Summing $Q_f + Q_t$
over all branches gives exactly
\[
  \sum_{\mathrm{branches}}(Q_f + Q_t) \;=\;
  Q_G^{\mathrm{sol}} - Q_D
  \;=\; 22.840 - 115.000 \;=\; -92.160~\MVAr.
\]
This is not a coincidence and not a violation of conservation; it is the
reactive analogue of the active-loss identity, but with a sign that can
go either way. Several branches inject more reactive power through their
shunt charging elements than they absorb in their series reactances, so
the per-branch reactive endpoint sum can be negative. In this case the
line charging dominates: the network as a whole is a net reactive-power
\emph{source}, supplying $92.160~\MVAr$ that the generators do not have
to provide. This is one of the reasons AC power flow must model
reactive power explicitly rather than treating it as a small correction
to a real-power-only calculation.

\subsection{Economic interpretation}

For readers approaching this from electricity markets rather than power
engineering, four observations on the same numbers are worth pulling out
explicitly.

\paragraph{Losses create a physical wedge that can enter prices.} The
$4.641~\MW$ of real losses is generation that has to be procured but
that no consumer pays for at a meter. In OPF and market models that
represent marginal losses, those losses enter the LMP decomposition as a
loss component alongside energy and congestion components. The
$4.641~\MW$ value here is the system-wide physical loss at this operating
point, not itself a price component. Losses are a real economic cost;
they just do not appear on the demand side of the energy-balance
constraint.

\paragraph{Reactive power is voltage support, not a second energy product.}
The $92.160~\MVAr$ that the network supplies from line charging is, in
effect, an uncontrolled capacitive injection in this model rather than a
generator dispatch decision. The $22.840~\MVAr$ from generators is a
solved operating-point quantity. In real systems generator reactive
output can consume capability and may be compensated through tariffs,
requirements, or ancillary-service mechanisms, but this fixture has no
generator capability curves or reactive-power cost data. The economic
lesson is therefore limited: reactive support is necessary and scarce in
operations, but these numbers alone do not imply a marginal cost.

\paragraph{Angle differences are the physical state behind DC flow.}
The DC power-flow approximation, which underlies many nodal market and
planning models, replaces the AC equations with a linearization in
which $P_{ij} \propto (\theta_i - \theta_j)/x_{ij}$. The angles are
primal state variables in that model, while LMPs are the dual variables
on the nodal balance constraints. Worked example: branch 8--2 carries
$163~\MW$ at an angle wedge of $5.56^\circ$, exactly tracking the
lossless formula. In a DC-OPF analogue, price separation would come from
binding constraints or modeled marginal losses, not from the mere
existence of a nonzero angle difference. The structural insight from the
AC calculation is therefore physical first: real power is driven by
angle differences; prices respond when those physical flows interact with
economic costs and constraints.

\paragraph{Bus 9 is the weakest-voltage load bus.}
Bus 9 has the lowest voltage ($0.9956~\pu$) and the most negative angle
($-3.99^\circ$). It is a large-load corner of the loop connected through
branches 8--9 and 9--4; it is not topologically isolated from the slack
injection point, but its local load and branch impedances make it the
lowest-voltage bus in this solution. In the LMP picture, those same
physical factors can contribute to price separation when they increase
marginal losses or make transmission constraints bind. A power-flow
solution alone does not determine which bus has the highest LMP; that
requires an OPF objective, generator costs, and active constraints.

\section{Takeaways}

\begin{enumerate}
  \item The IEEE 9-bus problem is a minimal but meaningful AC power-flow
  benchmark: it includes slack, PV, and PQ bus behavior in a small meshed
  network with three generator step-up transformers and a six-bus
  transmission loop.
  \item The unknown state consists of voltage angles at non-slack buses
  and voltage magnitudes at PQ buses. Slack-bus real and reactive power
  and PV-bus reactive power are solved outputs, not specified inputs.
  The slack bus absorbs the active-power residual from losses; reactive
  balance is shared through the solved reactive outputs of the slack and
  PV buses. Case-side $P_G$ at the slack and $Q_G$ at every generator
  should be read as initial values, not setpoints.
  \item The solved real-power generation exceeds total real load by
  $4.641~\MW$, which is the network real-power loss at this operating
  point.
  \item Voltage angle differences govern active-power flows through the
  near-lossless relation $P_{ij} \approx (|V_i||V_j|/x_{ij})
  \sin(\theta_i - \theta_j)$. Branch 8--2 reproduces its solved
  $163~\MW$ flow from this formula to three figures.
  \item Branch-flow signs must be read relative to the stored branch
  orientation. A negative endpoint power means power is delivered from
  the branch into that bus terminal rather than injected from the bus
  into the branch.
  \item Reactive-power behavior is qualitatively different from
  real-power loss accounting because line charging can inject reactive
  power. In this case the network is a net reactive-power source of
  $92.160~\MVAr$, and an interior PQ bus (bus 6) sits above the generator
  PV setpoint.
  \item The vocabulary has useful but limited market analogies:
  angle differences drive real-power exchange, real power is the
  commodity, reactive power is voltage support, losses are transaction
  costs, and the slack bus fixes the reference angle. AC power flow is
  the feasibility layer beneath nodal OPF and LMP calculations.
\end{enumerate}

\appendix

\section{Newton--Raphson Solution}\label{app:newton}

The non-slack bus active-power mismatches and the PQ-bus reactive-power
mismatches are
\[
  \Delta P_i = P_i^{\mathrm{spec}} - P_i(V,\theta), \quad i \notin \{\text{slack}\},
\]
\[
  \Delta Q_i = Q_i^{\mathrm{spec}} - Q_i(V,\theta), \quad i \in \{\text{PQ}\}.
\]
With $n_{\mathrm{PV}}$ PV buses, the unknowns are the $n - 1$ non-slack
voltage angles and the $n - n_{\mathrm{PV}} - 1$ PQ-bus voltage
magnitudes, giving a square nonlinear system. Let $J_{\mathrm{calc}}$ be
the Jacobian of the calculated injections,
\[
  J_{\mathrm{calc}} =
  \begin{bmatrix}
    \dfrac{\partial P}{\partial \theta} & \dfrac{\partial P}{\partial |V|} \\[6pt]
    \dfrac{\partial Q}{\partial \theta} & \dfrac{\partial Q}{\partial |V|}
  \end{bmatrix}
  \;=\;
  \begin{bmatrix} H & N \\ M & L \end{bmatrix},
\]
restricted to the rows and columns corresponding to the unknowns above.
Because the mismatch is defined as specified minus calculated injection,
the Newton step solves
\[
  J_{\mathrm{calc}}
  \begin{bmatrix} \Delta\theta \\ \Delta|V| \end{bmatrix}
  =
  \begin{bmatrix} \Delta P \\ \Delta Q \end{bmatrix},
\]
then adds the step to the current voltage angles and magnitudes. If one
instead defines the residual Jacobian directly, it is
$-J_{\mathrm{calc}}$ and the right-hand side changes sign as well.
Iteration continues until $\max_i (|\Delta P_i|, |\Delta Q_i|)$ falls
below a tolerance, typically $10^{-6}$ to $10^{-8}$ p.u. The MATPOWER
fixture used here does not record the iteration count or final mismatch,
so we cite only the converged state.

\end{document}
